### 2025-04-03 scripts/Paper-analysis.Rmd
link figure titles to code

# Maximum counts
d.reduced.median.plot = d.reduced.median %>% 
  group_by(PTU) %>%
  mutate(max_n = max(n, na.rm = TRUE),  # Calculate max 'n' for each 'PTU'
         normalized_n = n / max_n * 100) %>%  # Normalize 'n' by this max value
  ungroup() %>%  # Remove the grouping
  select(-max_n)  # Optionally remove the max_n column if not needed
# Plot the distribution
p.lines.expectation = ggplot(d.reduced.median.plot, aes(x, median*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.expectation, file='../results/6mer-deviation-expectation.pdf',
       width=10, height=6)
# Plot only first 25kb
ggsave(p.lines.expectation+scale_x_continuous(labels = function(x) x / 1000, limits=c(0,50000)), file='../results/6mer-deviation-expectation-0-50-kb.pdf',
       width=10, height=6)

@@

### Compare leading region to rest of plasmid

@@

# Order the PTUs by mean values of depletion for better visual display
mean.values = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean=mean(diff.density)*1000)
diff.leading.region.average$PTU = ordered(gsub("PTU-", "", diff.leading.region.average$PTU),
                                          levels=gsub("PTU-", "", mean.values$PTU[order(mean.values$mean)]))
p.leading.region.diff.density = ggplot(diff.leading.region.average, aes(as.factor(PTU), diff.density*1000))+ # depletion per-kb rather than per-bp
  ggbeeswarm::geom_quasirandom(alpha=0.8)+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0)+
  xlab("")+
  theme(axis.text=element_text(colour="black", size=18))+
  ylab("Palindrome depletion in leading region (per kb)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  ylab("")
p.leading.region.diff.density
# Based on this, remove two outliers
p.leading.region.diff.density = p.leading.region.diff.density+
  scale_y_continuous(breaks=seq(-8, 4, 2), limits=c(-8,5))
ggsave(p.leading.region.diff.density, file='../results/6mer-deviation-leading-region-vs-rest-of-plasmid-density.pdf',
       width=5.5, height=6)
```

Now we want to correlate the GC content difference of the first 5kb of the plasmid with the difference in palindrome density

@@

d.GC.palindrome.PTU = d.GC.palindrome %>% group_by(PTU) %>%
  summarise(GC=mean(GC),
            diff.GC=mean(diff.GC),
            density=mean(density),
            density.depletion=mean(density.depletion),
            n=length(window_size))
ggplot(d.GC.palindrome, aes(diff.GC, density.depletion, fill=PTU))+
  geom_point(shape=21, colour="black")+
  facet_wrap(~PTU)
p.gc.correlation = ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()+
  theme_bw()+
  xlab("Relative GC content at 5-10kb (%)")+
  ylab("Relative palindrome depletion at 5-10kb (per kb)")+
    xlim(c(-15, 10))+
  ylim(-13, 14)
p.gc.correlation.ptu =  ggplot(d.GC.palindrome.PTU, aes(diff.GC, density.depletion))+
  geom_point(aes(size=n, fill=PTU), shape=21, colour="black")+
  theme_bw()+
  ggrepel::geom_text_repel(aes(label=PTU), size=2, min.segment.length = 0)+
  guides(fill=FALSE)+
    xlab("Relative GC content (%)")+
  ylab("Relative palindrome depletion (per kb)")
pdf('../results/GC-correlation-depletion.pdf', width=8, height=4)
cowplot::plot_grid(p.gc.correlation+ggtitle("(a)"), p.gc.correlation.ptu+ggtitle("(b)"))
dev.off()
cor.test(d.GC.palindrome$diff.GC, d.GC.palindrome$density.depletion, method="spearman")
cor.test(d.GC.palindrome.PTU$diff.GC, d.GC.palindrome.PTU$density.depletion)

@@

Similar to Samuel et al.'s display of anti-defence gene density, average over all conjugative plasmids.

```{r}
overall.median = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=median(diff_density),
            n=length(diff_density))
p.overall.average = ggplot(overall.median, aes(x, y*1000))+
  geom_line()+
  scale_y_continuous(breaks=seq(-10, 0, 2), limits = c(-9, 0))+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0, linetype='dashed')+
  ylab("Palindrome depletion (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.overall.average
# Save
ggsave(p.overall.average, file='../results/6mer-deviation-overall.pdf', 
       width=3, height=3)
```
Also do this for GC content.

@@

Test for a particular PTU.

```{r}

@@

p.depletion.targets = ggplot(results, aes(median.score, as.factor(PTU)))+
  geom_vline(xintercept = 0, size=1)+
  geom_bar(stat="identity", position="dodge")+
  facet_wrap(~k, scales="free_x")+
  ylab("")+
  xlab("Median depletion score of within-range RM targets\n(leading vs. lagging, 10kb)")+
   theme_bw()+
  theme(panel.grid = element_blank())+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))
ggsave(p.depletion.targets, file='../results/figure-depletion-within-range-targets.pdf',
       width=10, height=6)

>>>>

# Maximum counts
d.reduced.median.plot = d.reduced.median %>% 
  group_by(PTU) %>%
  mutate(max_n = max(n, na.rm = TRUE),  # Calculate max 'n' for each 'PTU'
         normalized_n = n / max_n * 100) %>%  # Normalize 'n' by this max value
  ungroup() %>%  # Remove the grouping
  select(-max_n)  # Optionally remove the max_n column if not needed
# Plot the distribution
p.lines.expectation = ggplot(d.reduced.median.plot, aes(x, median*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.expectation, file='../results/6mer-deviation-expectation.pdf',
       width=10, height=6)
# Plot only first 50kb
ggsave(p.lines.expectation+scale_x_continuous(labels = function(x) x / 1000, limits=c(0,50000)), file='../results/6mer-deviation-expectation-0-50-kb.pdf',
       width=10, height=6)

@@

### Figure 2

Figure 2. RM target depletion in the leading region of conjugative plasmids.

```{r}
overall.median = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=median(diff_density),
            n=length(diff_density))
p.overall.average = ggplot(overall.median, aes(x, y*1000))+
  geom_line()+
  scale_y_continuous(breaks=seq(-10, 0, 2), limits = c(-9, 0))+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0, linetype='dashed')+
  ylab("Palindrome depletion (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.overall.average
# Save
ggsave(p.overall.average, file='../results/Figure-2-6mer-deviation-overall.pdf', 
       width=3, height=3)
```

@@

# Order the PTUs by mean values of depletion for better visual display
mean.values = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean=mean(diff.density)*1000)
diff.leading.region.average$PTU = ordered(gsub("PTU-", "", diff.leading.region.average$PTU),
                                          levels=gsub("PTU-", "", mean.values$PTU[order(mean.values$mean)]))
p.leading.region.diff.density = ggplot(diff.leading.region.average, aes(as.factor(PTU), diff.density*1000))+ # depletion per-kb rather than per-bp
  ggbeeswarm::geom_quasirandom(alpha=0.8)+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0)+
  xlab("")+
  theme(axis.text=element_text(colour="black", size=18))+
  ylab("Palindrome depletion in leading region (per kb)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  ylab("")
p.leading.region.diff.density
# Based on this, remove two outliers
p.leading.region.diff.density = p.leading.region.diff.density+
  scale_y_continuous(breaks=seq(-8, 4, 2), limits=c(-8,5))
ggsave(p.leading.region.diff.density, file='../results/Figure-2-6mer-deviation-leading-region-vs-rest-of-plasmid-density.pdf',
       width=5.5, height=6)
```

### Figure 3

Figure 3. GC-rich leading regions are more depleted in 6-bp palindromes. Values are calculated for the 5-10kb section of the leading region relative to the mean average value over the whole length of the plasmid. Spearman’s correlations are given as inset values. The relationship between relative GC content and palindrome depletion was still significant in a linear model controlling for PTU (depletion ~ GC*PTU).

@@

d.GC.palindrome.PTU = d.GC.palindrome %>% group_by(PTU) %>%
  summarise(GC=mean(GC),
            diff.GC=mean(diff.GC),
            density=mean(density),
            density.depletion=mean(density.depletion),
            n=length(window_size))
ggplot(d.GC.palindrome, aes(diff.GC, density.depletion, fill=PTU))+
  geom_point(shape=21, colour="black")+
  facet_wrap(~PTU)
p.gc.correlation = ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()+
  theme_bw()+
  xlab("Relative GC content at 5-10kb (%)")+
  ylab("Relative palindrome depletion at 5-10kb (per kb)")+
    xlim(c(-15, 10))+
  ylim(-13, 14)
p.gc.correlation.ptu =  ggplot(d.GC.palindrome.PTU, aes(diff.GC, density.depletion))+
  geom_point(aes(size=n, fill=PTU), shape=21, colour="black")+
  theme_bw()+
  ggrepel::geom_text_repel(aes(label=PTU), size=2, min.segment.length = 0)+
  guides(fill=FALSE)+
    xlab("Relative GC content (%)")+
  ylab("Relative palindrome depletion (per kb)")
pdf('../results/Figure-3-GC-correlation-depletion.pdf', width=8, height=4)
cowplot::plot_grid(p.gc.correlation+ggtitle("(a)"), p.gc.correlation.ptu+ggtitle("(b)"))
dev.off()
cor.test(d.GC.palindrome$diff.GC, d.GC.palindrome$density.depletion, method="spearman")
cor.test(d.GC.palindrome.PTU$diff.GC, d.GC.palindrome.PTU$density.depletion)

@@

Also do this for GC content.

@@

### Figure 4

Figure 4: Average depletion of within-range RM targets in the leading vs. lagging regions for conjugative PTUs. 
```{r}

@@

p.depletion.targets = ggplot(results, aes(median.score, as.factor(PTU)))+
  geom_vline(xintercept = 0, size=1)+
  geom_bar(stat="identity", position="dodge")+
  facet_wrap(~k, scales="free_x")+
  ylab("")+
  xlab("Median depletion score of within-range RM targets\n(leading vs. lagging, 10kb)")+
   theme_bw()+
  theme(panel.grid = element_blank())+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))
ggsave(p.depletion.targets, file='../results/Figure-4-figure-depletion-within-range-targets.pdf',
       width=10, height=6)
